preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-03 Manuscript/manuscript.Rmd
More edits

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. First, the data matrix $A_{N \times (p+1)}$ was constructed by combining the binned genome from each of the $n-1$ autosomal chromosomes ($1,2,\cdots, i-1,i+1,\cdots,22$) as well as a particular feature engineering procedure. Re-sampling was then performed on $A$. An RF classifier was then trained on $A$ using 3-fold cross-validation to tune for the number of annotations, *p*, to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. To evaluate model performance, the $i^{th}$ chromosome was used for testing. The performance of each combination of model inputs was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

$$BA = \dfrac{1}{2}\left( \dfrac{TP}{TP+FN} + \dfrac{TN}{TN+FP} \right)$$

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we implemented an optimally built RF model using the framework described above. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements).

First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, *DBSCAN* (version 1.1-5) was applied to the matrix of pairwise genomic distances between base pairs with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The eps parameter was assigned as the respective data resolution. The heuristic being that density-reachable base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base pair among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. Thusly, the corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base pair coordinate within each clustered PTBR.

@@

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix and plotHeatmap tools offered by deepTools version 2.0 [@ramirez2016deeptools2] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we assessed the hypothesis that TADs and chromatin loops are conserved  [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa] by comparing predicted and called boundaries between the GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) and evaluated using the Jaccard index defined as

>>>>

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. First, the data matrix $A_{N \times (p+1)}$ was constructed by combining the binned genome from each of the $n-1$ autosomal chromosomes ($1,2,\cdots, i-1,i+1,\cdots,22$) [??? Define N and p asap introducing. What is i? also define immediately. Adjust this whole paragraph accordingly. There's a term "hold-out" - isn't it how you should define it?] as well as a particular feature engineering procedure. Re-sampling was then performed on $A$. An RF classifier was then trained on $A$ using 3-fold cross-validation to tune for the number of annotations, *p*, to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. To evaluate model performance, the $i^{th}$ chromosome was used for testing. The performance of each combination of model inputs was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

$$BA = \dfrac{1}{2}\left( sensitivity + specificity \right) = \dfrac{1}{2}\left( \dfrac{TP}{TP+FN} + \dfrac{TN}{TN+FP} \right)$$

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we implemented an optimally built RF model using the framework described above. That is, a random forest model was built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements).

[??? There is a big gap there. Describe the construction of dependent vector Y and predictor matrix _for each base_]

First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, *DBSCAN* [??? We need to define what to italicize. Check below, make consistent throughout the manuscript] (version 1.1-5) was applied to the matrix of pairwise genomic distances between base pairs with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The *eps* parameter was assigned as the respective data resolution. The heuristic being that density-reachable base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding *MinPts* parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base pair among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base pair coordinate within each clustered PTBR.

@@

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix [??? Again, I expected italics, but decide and correct] and plotHeatmap tools offered by deepTools version 2.0 [@ramirez2016deeptools2] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we assessed the hypothesis that TADs and chromatin loops are conserved  [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa] by comparing predicted and called boundaries between the GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) and evaluated using the Jaccard index defined as
